Exploratory_Analysis_Josh_Tobey

### 2025-12-10 Exploratory_Analysis_Josh_Tobey.Rmd
Updated file with gam and plot against fitted and actuals.

```{r}
time_lag_1 <- 
time_lag_2 <- 
time_lag_3 <-

weekly_log_spec <- logistic_reg()
weekly_log_fit <- weekly_log_spec |>
  fit(Direction ~ Lag1 + 
      Lag2, 
      family = "binomial", 
      data = Weekly)
weekly_log_fit |>
  pluck("fit") |>
  summary()

@@

```{r}
cross_validation <- initial_split(exploratory_df, prop = 0.6)
exploratory_train <- training(cross_validation)
exploratory_test <- testing(cross_validation)
#OPTION 1:
linear_recipe <- recipe(`pH SU` ~ `Temperature deg C`, data = exploratory_df)
quad_recipe <- linear_recipe |> step_mutate(`Temperature deg C_2` = `Temperature deg C`^2)
cubic_recipe <- quad_recipe |> step_mutate(`Temperature deg C_3` = `Temperature deg C`^3)
quart_recipe <- cubic_recipe |> step_mutate(`Temperature deg C_4` = `Temperature deg C`^4)
#OPTION 2:
linear_recipe <- recipe(`pH SU` ~ ., data = exploratory_df)
quad_recipe <- linear_recipe |> step_poly(all_numeric_predictors(), degree = 2)
cubic_recipe <- quad_recipe |> step_poly(all_numeric_predictors(), degree = 3)
quart_recipe <- cubic_recipe |> step_poly(all_numeric_predictors(), degree = 4)

@@

```
```{r}
RWAR_dosing_lag_incdec <- RWAR_dosing_lag_incdec %>%
  select(-c(FeCl_change, Polymer_change, Timestamp))
head(RWAR_dosing_lag_incdec)
```

```{r}
library(mgcv)

lm_spec <- linear_reg()

gam_formula <- Ferric.chloride..mg.L ~ s(Cationic.Polimer..mg.L) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) + s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) + s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) + s(Turb_lag1) + s(Turb_lag2)

gam_fit <- gam(gam_formula, data = RWAR_dosing_lag_incdec)
gam_rec <- recipe(Ferric.chloride..mg.L ~ ., data = RWAR_dosing_lag_incdec) %>%
  step_ns(Cationic.Polimer..mg.L, deg_free = 10) %>%
  step_ns(mean_Temp, deg_free = 10) %>%
  step_ns(mean_pH, deg_free = 10) %>%
  step_ns(mean_Conductivity, deg_free = 10) %>%
  step_ns(mean_Turbidity, deg_free = 10) %>%
  step_ns(Temp_lag1, deg_free = 10) %>%
  step_ns(Temp_lag2, deg_free = 10) %>%
  step_ns(pH_lag1, deg_free = 10) %>%
  step_ns(pH_lag2, deg_free = 10) %>%
  step_ns(Cond_lag1, deg_free = 10) %>%
  step_ns(Cond_lag2, deg_free = 10) %>%
  step_ns(Turb_lag1, deg_free = 10) %>%
  step_ns(Turb_lag2, deg_free = 10)

gam_wf <- workflow() |>
  add_model(lm_spec) |>
  add_recipe(gam_rec)

gam_fit <- gam_wf |>
  fit(data = RWAR_dosing_lag_incdec)

gam_fit |> tidy()
gam_fit |> glance()

dat_grid <- tibble(expand.grid(Cationic.Polimer..mg.L = seq(min(RWAR_dosing_lag_incdec$Cationic.Polimer..mg.L), 
                               max(RWAR_dosing_lag_incdec$Cationic.Polimer..mg.L)),
                               mean_Temp = seq(min(RWAR_dosing_lag_incdec$mean_Temp), max(RWAR_dosing_lag_incdec$mean_Temp)),
                               mean_pH = seq(min(RWAR_dosing_lag_incdec$mean_pH), max(RWAR_dosing_lag_incdec$mean_pH)),
                               mean_Conductivity = seq(min(RWAR_dosing_lag_incdec$mean_Conductivity), max(RWAR_dosing_lag_incdec$mean_Conductivity)),
                               mean_Turbidity = seq(min(RWAR_dosing_lag_incdec$mean_Turbidity), max(RWAR_dosing_lag_incdec$mean_Turbidity)),
                               Temp_lag1 = seq(min(RWAR_dosing_lag_incdec$Temp_lag1), max(RWAR_dosing_lag_incdec$Temp_lag1)),
                               Temp_lag2 = seq(min(RWAR_dosing_lag_incdec$Temp_lag2), max(RWAR_dosing_lag_incdec$Temp_lag2)),
                               pH_lag1 = seq(min(RWAR_dosing_lag_incdec$pH_lag1), max(RWAR_dosing_lag_incdec$pH_lag1)),
                               pH_lag2 = seq(min(RWAR_dosing_lag_incdec$pH_lag2), max(RWAR_dosing_lag_incdec$pH_lag2)),
                               Cond_lag1 = seq(min(RWAR_dosing_lag_incdec$Cond_lag1), max(RWAR_dosing_lag_incdec$Cond_lag1)),
                               Cond_lag2 = seq(min(RWAR_dosing_lag_incdec$Cond_lag2), max(RWAR_dosing_lag_incdec$Cond_lag2)),
                               Turb_lag1 = seq(min(RWAR_dosing_lag_incdec$Turb_lag1), max(RWAR_dosing_lag_incdec$Turb_lag1)),
                               Turb_lag2 = seq(min(RWAR_dosing_lag_incdec$Turb_lag2), max(RWAR_dosing_lag_incdec$Turb_lag2))
                               ))
bind_cols(
  augment(gam_fit, new_data = dat_grid),
  predict(gam_fit, new_data = dat_grid, type = "conf_int")
) -> dose_pred_gam

for(pred in c("age", "year", "education")) {
  pred_dat <- dose_pred_gam[, c(pred, ".pred", ".pred_lower", ".pred_upper")] |>
    group_by_at(pred) |>
    summarise(.pred = mean(.pred), .pred_lower = mean(.pred_lower), .pred_upper = mean(.pred_upper))
  dose_dat <- RWAR_dosing_lag_incdec[, c(pred, "Ferric.chloride..mg.L")]

  if(is.numeric(dose_dat |> pull(pred))) {
    p <- dose_dat |>
      ggplot() +
      geom_point(aes_string(pred, "wage"), alpha = .3) +
      geom_line(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
      geom_line(aes_string(pred, ".pred_lower"), data = pred_dat, colour = "darkgreen", lty = 2) +
      geom_line(aes_string(pred, ".pred_upper"), data = pred_dat, colour = "darkgreen", lty = 2)
  } else {
    p <- dose_dat |>
      ggplot() +
      geom_jitter(aes_string(pred, "wage"), alpha = .3) +
      geom_point(aes_string(pred, ".pred"), data = pred_dat, colour = "darkgreen") +
      geom_errorbar(aes_string(pred, ymin = ".pred_lower", ymax = ".pred_upper"), data = pred_dat, colour = "darkgreen")
  }
  print(p)
}

@@

>>>>

```{r, eval = FALSE}
weekly_log_spec <- logistic_reg()
weekly_log_fit <- weekly_log_spec |>
  fit(Direction ~ Lag1 + 
      Lag2, 
      family = "binomial", 
      data = Weekly)
weekly_log_fit |>
  pluck("fit") |>
  summary()

@@

```{r, eval = FALSE}
cross_validation <- initial_split(exploratory_df, prop = 0.6)
exploratory_train <- training(cross_validation)
exploratory_test <- testing(cross_validation)
#OPTION 1:
linear_recipe <- recipe(`pH SU` ~ `Temperature deg C`, data = exploratory_df)
quad_recipe <- linear_recipe |> step_mutate(`Temperature deg C_2` = `Temperature deg C`^2)
cubic_recipe <- quad_recipe |> step_mutate(`Temperature deg C_3` = `Temperature deg C`^3)
quart_recipe <- cubic_recipe |> step_mutate(`Temperature deg C_4` = `Temperature deg C`^4)
#OPTION 2:
linear_recipe <- recipe(`pH SU` ~ ., data = exploratory_df)
quad_recipe <- linear_recipe |> step_poly(all_numeric_predictors(), degree = 2)
cubic_recipe <- quad_recipe |> step_poly(all_numeric_predictors(), degree = 3)
quart_recipe <- cubic_recipe |> step_poly(all_numeric_predictors(), degree = 4)

@@

```
```{r}
RWAR_dosing_lag_incdec <- RWAR_dosing_lag_incdec %>%
  select(-c(FeCl_change, Polymer_change))
head(RWAR_dosing_lag_incdec)
```

```{r}
library(mgcv)
library(parsnip)
df_full <- RWAR_dosing_lag_incdec
df_train <- df_full |> filter(year(Timestamp) <= 2019)
df_test  <- df_full |> filter(year(Timestamp) == 2020)

gam_spec <- gen_additive_mod() |> 
  set_engine("mgcv") |> 
  set_mode("regression")
gam_rec <- recipe(FeCl_dose ~ Polymer_dose + mean_Temp + mean_pH + mean_Conductivity +
                 mean_Turbidity + Temp_lag1 + Temp_lag2 +
                 pH_lag1 + pH_lag2 + Cond_lag1 + Cond_lag2 +
                 Turb_lag1 + Turb_lag2, data = df_train)

gam_wf <- workflow() |>
  add_model(gam_spec) |>
  add_recipe(gam_rec)

gam_fit <- gam(
  FeCl_dose ~ s(Polymer_dose) + s(mean_Temp) + s(mean_pH) + s(mean_Conductivity) +
               s(mean_Turbidity) + s(Temp_lag1) + s(Temp_lag2) +
               s(pH_lag1) + s(pH_lag2) + s(Cond_lag1) + s(Cond_lag2) +
               s(Turb_lag1) + s(Turb_lag2),
  data = df_train
)

gam_fit |> tidy()
gam_fit |> glance()
```
```{r}
pred_df <- df_test %>% 
  mutate(pred = predict(gam_fit, newdata = df_test))
rmse(pred_df, truth = FeCl_dose, estimate = pred)
#The test data shows that

@@

```{r}

dat_grid_dose <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Polymer_dose = seq(min(df_train$Polymer_dose),
                                      max(df_train$Polymer_dose),
                                      length.out = 100))
dat_grid_temp <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_Temp = seq(min(df_train$mean_Temp), 
                         max(df_train$mean_Temp),
                                      length.out = 100))
dat_grid_pH <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_pH = seq(min(df_train$mean_pH), max(df_train$mean_pH),
                                      length.out = 100))
dat_grid_cond <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_Conductivity = seq(min(df_train$mean_Conductivity), max(df_train$mean_Conductivity),
                                      length.out = 100))
dat_grid_turb <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(mean_Turbidity = seq(min(df_train$mean_Turbidity), max(df_train$mean_Turbidity),
                                      length.out = 100))
dat_grid_temp_lag1 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Temp_lag1 = seq(min(df_train$Temp_lag1), max(df_train$Temp_lag1),
                                      length.out = 100))
dat_grid_temp_lag2 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Temp_lag2 = seq(min(df_train$Temp_lag2), max(df_train$Temp_lag2),
                                      length.out = 100))
dat_grid_pH_lag1 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(pH_lag1 = seq(min(df_train$pH_lag1), max(df_train$pH_lag1),
                                      length.out = 100))
dat_grid_pH_lag2 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(pH_lag2 = seq(min(df_train$pH_lag2), max(df_train$pH_lag2),
                                      length.out = 100))
dat_grid_cond_lag1 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Cond_lag1 = seq(min(df_train$Cond_lag1), max(df_train$Cond_lag1),
                                      length.out = 100))
dat_grid_cond_lag2 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Cond_lag2 = seq(min(df_train$Cond_lag2), max(df_train$Cond_lag2),
                                      length.out = 100))
dat_grid_turb_lag1 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Turb_lag1 = seq(min(df_train$Turb_lag1), max(df_train$Turb_lag1),
                                      length.out = 100))
dat_grid_turb_lag2 <- df_train |> 
  summarize(
    across(everything(), mean, na.rm = TRUE)
  ) |> 
  slice(rep(1, 100)) |> 
  mutate(Turb_lag2 = seq(min(df_train$Turb_lag2), max(df_train$Turb_lag2),
                                      length.out = 100))

pred1 <- predict(gam_fit, newdata = dat_grid_dose, type = "response", se.fit = TRUE)
dose_pred <- dat_grid_dose |> mutate(
  .fitted = pred1$fit,
  .lower = pred1$fit - 2 * pred1$se.fit,
  .upper = pred1$fit + 2 * pred1$se.fit
  )
pred2 <- predict(gam_fit, newdata = dat_grid_temp, type = "response", se.fit = TRUE)
temp_pred <- dat_grid_temp |> mutate(
  .fitted = pred2$fit,
  .lower = pred2$fit - 2 * pred2$se.fit,
  .upper = pred2$fit + 2 * pred2$se.fit
)
pred3 <- predict(gam_fit, newdata = dat_grid_pH, type = "response", se.fit = TRUE)
pH_pred <- dat_grid_pH |> mutate(
  .fitted = pred3$fit,
  .lower = pred3$fit - 2 * pred3$se.fit,
  .upper = pred3$fit + 2 * pred3$se.fit
)
pred4 <- predict(gam_fit, newdata = dat_grid_cond, type = "response", se.fit = TRUE)
cond_pred <- dat_grid_cond |> mutate(
  .fitted = pred4$fit,
  .lower = pred4$fit - 2 * pred4$se.fit,
  .upper = pred4$fit + 2 * pred4$se.fit
)
pred5 <- predict(gam_fit, newdata = dat_grid_turb, type = "response", se.fit = TRUE)
turb_pred <- dat_grid_turb |> mutate(
  .fitted = pred5$fit,
  .lower = pred5$fit - 2 * pred5$se.fit,
  .upper = pred5$fit + 2 * pred5$se.fit
)
pred6 <- predict(gam_fit, newdata = dat_grid_temp_lag1, type = "response", se.fit = TRUE)
tempLag1_pred <- dat_grid_temp_lag1 |> mutate(
  .fitted = pred6$fit,
  .lower = pred6$fit - 2 * pred6$se.fit,
  .upper = pred6$fit + 2 * pred6$se.fit
)
pred7 <- predict(gam_fit, newdata = dat_grid_temp_lag2, type = "response", se.fit = TRUE)
tempLag2_pred <- dat_grid_temp_lag2 |> mutate(
  .fitted = pred7$fit,
  .lower = pred7$fit - 2 * pred7$se.fit,
  .upper = pred7$fit + 2 * pred7$se.fit
)
pred8 <- predict(gam_fit, newdata = dat_grid_pH_lag1, type = "response", se.fit = TRUE)
pHLag1_pred <- dat_grid_pH_lag1 |> mutate(
  .fitted = pred8$fit,
  .lower = pred8$fit - 2 * pred8$se.fit,
  .upper = pred8$fit + 2 * pred8$se.fit
)
pred9 <- predict(gam_fit, newdata = dat_grid_pH_lag2, type = "response", se.fit = TRUE)
pHLag2_pred <- dat_grid_pH_lag2 |> mutate(
  .fitted = pred9$fit,
  .lower = pred9$fit - 2 * pred9$se.fit,
  .upper = pred9$fit + 2 * pred9$se.fit
)
pred10 <- predict(gam_fit, newdata = dat_grid_cond_lag1, type = "response", se.fit = TRUE)
condLag1_pred <- dat_grid_cond_lag1 |> mutate(
  .fitted = pred10$fit,
  .lower = pred10$fit - 2 * pred10$se.fit,
  .upper = pred10$fit + 2 * pred10$se.fit
)
pred11 <- predict(gam_fit, newdata = dat_grid_cond_lag2, type = "response", se.fit = TRUE)
condLag2_pred <- dat_grid_cond_lag2 |> mutate(
  .fitted = pred11$fit,
  .lower = pred11$fit - 2 * pred11$se.fit,
  .upper = pred11$fit + 2 * pred11$se.fit
)
pred12 <- predict(gam_fit, newdata = dat_grid_turb_lag1, type = "response", se.fit = TRUE)
turbLag1_pred <- dat_grid_turb_lag1 |> mutate(
  .fitted = pred12$fit,
  .lower = pred12$fit - 2 * pred12$se.fit,
  .upper = pred12$fit + 2 * pred12$se.fit
)
pred13 <- predict(gam_fit, newdata = dat_grid_turb_lag2, type = "response", se.fit = TRUE)
turbLag2_pred <- dat_grid_turb_lag2 |> mutate(
  .fitted = pred13$fit,
  .lower = pred13$fit - 2 * pred13$se.fit,
  .upper = pred13$fit + 2 * pred13$se.fit
)

```{r}
summary(gam_fit)
```

```{r}
ggplot(pred_df) +
  geom_point(aes(x = FeCl_dose, y = pred), alpha = 0.3) +
  geom_abline(slope = 1, 
              intercept = coef(gam_fit)[1],
              color = 'red',
              lwd = 1)

```
```{r}
ggplot(df_test) +
  geom_line(aes(y = FeCl_dose), color = 'orange')

The GAM model shows poor performance when making prediction of the dose requirements for water treatment.
